---
title: Lipidomics report
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
output_file: lip
params: 
  metadata_file: ../data/metadata.txt
  input_file: ../data/alldata.lipidomics_summary_test_data.tsv
  #input_file: x
  meta_sample_name: SampleName
  meta_condition_type: Type
  meta_rep: Rep
  lipid_name_col: X1
  paired: TRUE

---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(tidyverse)
options(knitr.duplicate.label = "allow")
source("heatmap_processing.R")
source("stats.R")
```

```{r import_data}
metadata <- readr::read_tsv(params$metadata_file)
all_data <- readr::read_tsv(params$input_file)

all_data <- dplyr::rename(
  all_data, lipid_name := tidyselect::all_of(params$lipid_name_col)
)
```

```{r process_data}
meta_info <- metadata %>%
  dplyr::rename(sample_name = tidyselect::all_of(params$meta_sample_name)) %>%
  dplyr::rename(condition = tidyselect::all_of(params$meta_condition_type)) %>%
  dplyr::mutate(condition = stringr::str_squish(condition)) #we've had double spaces that messes things up

if(params$paired) {
  meta_type <- meta_info %>%
    dplyr::rename(rep = tidyselect::all_of(params$meta_rep)) %>%
    dplyr::select(sample_name, condition, rep) %>%
    dplyr::mutate(rep = forcats::as_factor(rep))
} else {
  meta_type <- meta_info %>%
    dplyr::select(sample_name, condition)
}

tidy_data <- all_data %>%
  tidyr::pivot_longer(cols = -lipid_name, names_to = "sample_name") %>%
  tidyr::separate(lipid_name, sep = " ", into = c("class", NA), remove = FALSE) %>%
  left_join(meta_type) %>%
  dplyr::mutate(log2_value = log2(value)) %>%
  dplyr::mutate(log2_value = if_else(log2_value == "-Inf", 0, log2_value)) # we don't want -Inf values

exp_conditions <- unique(pull(tidy_data, condition))
```

# {.tabset .tabset-pills}

## Summary plots {.active}

### Bar plots for lipid classes
These plots show linear mean values for each condition within each lipid subclass. 
The errorbars show the standard error of the mean.  
More information can be found in the other tabs in the report.
  
```{r}
# for if there are loads and we want a link to the plot further down the page
lipid_classes <- unique(tidy_data$class)
class_link <- function(class){
  paste0("<a href=", "\"", paste0("#", tolower(class)), "\">", class, "</a>")
}  
```

#### `r sapply(lipid_classes, class_link)`

```{r, results = "asis"}
res <- lapply(lipid_classes, function(lipid_class) {
  knitr::knit_child(
    "bar_plots.Rmd", envir = environment(), quiet = TRUE
  )
})
cat(unlist(res), sep = '\n')
```


## Summary table

```{r calculate_stats_individual_lipids}
# get mean and standard deviation
mean_stdev <- tidy_data %>%
  dplyr::group_by(lipid_name, condition) %>%
  dplyr::summarise(
    mean = mean(value), 
    stdev = sd(value),
    se = sd(value)/sqrt(dplyr::n())
  ) %>%
  ungroup()
  
fc <- get_fold_change(mean_stdev) 

# for calculating the stats per individual lipid - functions are in stats.R 
stats_summary <- mean_stdev %>%
  group_by(lipid_name) %>%
  mutate(sd_ratio = get_sd_ratio(cur_data())) %>%
  mutate(valid_ratio = get_ratio_validity(cur_data())) %>%
  ungroup() %>%
  left_join(fc) %>%
  tidyr::separate(lipid_name, sep = " ", into = c("class", NA), remove = FALSE)

long_stats_pvals <- do_stats(tidy_data, stats_summary, params$paired) 

# multiple testing correction - by class for now.
long_stats_pvals <- long_stats_pvals %>%
  select(lipid_name, class, p_val) %>%
  distinct() %>%
  group_by(class) %>%
  mutate(adj_pval = p.adjust(p_val, method = "BH")) %>%
  ungroup() %>%
  left_join(long_stats_pvals)

stats_pvals <- long_stats_pvals %>%
  select(lipid_name, test_type, p_val, adj_pval) %>%
  distinct() %>%
  right_join(stats_summary)

```

```{r}
summary_table_data <- stats_pvals %>%
  select(-contains("ratio"), -stdev, -se) %>%
  dplyr::relocate(class) %>%
  pivot_wider(names_from = condition, names_prefix = "mean_", values_from = mean) 

log2_fc <- summary_table_data %>%
  select(starts_with("FC")) %>%
  log2() %>%
  rename_with(~ paste0("log2_", .x))

summary_table_data <- cbind(summary_table_data, log2_fc)

cols_to_round <- summary_table_data %>%
  dplyr::select(all_of(c(contains("mean"), contains("FC")))) %>%
  colnames()

pvals_to_round <- summary_table_data %>%
  dplyr::select(all_of(c(contains("p_val"), contains("pval")))) %>%
  colnames()

summary_table_data %>%
  DT::datatable(rownames = FALSE, options = list(pageLength = 15, dom = "ftlip")) %>%
  DT::formatRound(digits = 3, cols_to_round) %>%
  DT::formatRound(digits = 5, pvals_to_round)
  
```

## Metadata

<div class = "col-md-6">
```{r}
DT::datatable(meta_type, rownames = FALSE, options = list(pageLength = 15, dom = "tp"))
```
</div>


## Individual lipids

#### Scroll down to browse plots for individual lipids or click on the name in the table below to navigate directly to the plot.

```{r}
# function to create a link in this format '<a href="#pc-280">PC 28:0</a>'
get_link <- function(lipid){
  temp <- gsub(" ", "-", x = tolower(lipid))
  part1 <- paste0("#", gsub(":", "", temp))
  paste0("<a href=", "\"", part1, "\">", lipid, "</a>")
}

summary_table_data %>%
  dplyr::mutate(lipid_name = get_link(lipid_name)) %>%
  #select(-fold_change) %>%
  DT::datatable(
    rownames = FALSE, 
    options = list(pageLength = 5, dom = "tlip"), 
    filter = list(position = "top"),
    escape = FALSE) %>%
  #DT::formatRound(5:7, digits = 3)
  DT::formatRound(digits = 3, cols_to_round)
  
```

```{r, results = "asis"}
res <- lapply(unique(tidy_data$lipid_name)[1:3], function(lipid_to_plot) { # to speed up tests
#res <- lapply(unique(tidy_data$lipid_name), function(lipid_to_plot) {
  knitr::knit_child(
    "child_plots.Rmd", envir = environment(), quiet = TRUE
  )
})
cat(unlist(res), sep = '\n')
```


## Heatmap

```{r, heatmap_processing}
#matrix_data <- remove_0_variance(all_data)

var_above0 <- all_data %>%
  pivot_longer(-lipid_name) %>%
  group_by(lipid_name) %>%
  summarise(variance = var(value)) %>%
  filter(variance > 0) %>%
  select(lipid_name) %>%
  left_join(all_data)

heatmap_data <- tibble::column_to_rownames(var_above0, "lipid_name")
 
```  

```{r, plot_heatmap}
classes <- tidy_data %>%
  select(lipid_name, class) %>%
  distinct()

# row names with their class
df_lipids <- var_above0 %>%
  select(lipid_name) %>%
  left_join(classes) 

row_annotation <- tibble::column_to_rownames(df_lipids, "lipid_name")

col_annotation <- enframe(colnames(heatmap_data), name = NULL, value = "sample_name") %>%
  left_join(meta_type) %>%
  tibble::column_to_rownames("sample_name")

# Lipid classes need further sorting to create custom row annotations where only
# the class name is displayed rather than each individual lipid.
lipid_summary <- get_lipid_summary(df_lipids)
lipid_labels <- create_lipid_class_labels(lipid_summary, nrow(df_lipids))
lipid_colours <- create_lipid_colours(
  class_names = dplyr::pull(lipid_summary, class)
)
group_colours <- stats::setNames(c("blue", "green"), exp_conditions)

annot_colours <- list(
  value = group_colours,
  class = lipid_colours
)
```

### Heatmap ordered by lipid subclass

```{r, fig.height = 8, fig.width = 10}
pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  cluster_rows = FALSE,
  labels_row = lipid_labels,
  fontsize_row = 8,
  annotation_colors = annot_colours#,
  #annotation_legend = FALSE
)
```

<br>
<br>

### With row clustering

```{r, fig.height = 8, fig.width = 10}
pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  annotation_colors = annot_colours
)
```

