
```{r}
dodge_width <- 0.9

# #datasets for individual lipids
# stats_pvals
# t_test_res
# mean_log2
# 
# # class dataset
# bar_data_linear_median # this is used for abundance level in the class plots

# the mean here is the mean of the 3 (or no of replicates) total values, not the overall mean
totals_data <- tidy_data %>%
  group_by(class, condition, sample_name) %>%
  summarise(sample_total = sum(value)) %>%
  ungroup() %>%
  group_by(class, condition) %>%
  dplyr::summarise(
    mean_total = mean(sample_total),
    se_total = sd(sample_total)/sqrt(dplyr::n()),
    sum_total = sum(sample_total)
  ) %>%
  ungroup() %>%
  group_by(condition) %>%
  mutate(percent = (mean_total/sum(mean_total))*100) %>%
  ungroup() %>%
  group_by(class) %>%
  mutate(max_percent = max(percent)) %>%
  mutate(max_mean_total = max(mean_total)) %>%
  ungroup() %>%
  mutate(pie_class = if_else(max_percent > 1, class, "other"))

#conditions <- unique(pull(tidy_data, condition))
conditions <- levels(tidy_data$condition)

mean_log2_class <- tidy_data %>%
  dplyr::group_by(class, condition) %>%#, sample_name) %>%
  dplyr::summarise(
    mean = mean(log2_value),
  ) %>%
  ungroup() %>%
  pivot_wider(names_from = condition, values_from = mean) %>%
  mutate(ratio = .data[[conditions[2]]] - .data[[conditions[1]]])
  #mutate(ratio = .data[[exp_conditions[1]]] - .data[[exp_conditions[2]]])
```

```{r}
t_test_by_class <- tidy_data %>%
  group_by(class) %>%
  select(class, sample_name, condition, log2_value) %>%
  rstatix::t_test(log2_value ~ condition, paired = params$paired) %>%
  ungroup() %>%
  mutate(padj = p.adjust(p, method="BH")) %>%
  ungroup() %>%
  mutate(tadj = qt(padj/2, df)) %>%
  left_join(mean_log2_class) %>%
  mutate(SE = abs(ratio/tadj), lower.limit=ratio-1.96*SE, upper.limit=ratio+1.96*SE) 

```

<hr class="style2">  

### Summary of total amounts across all samples

The pie charts show the proportion of lipid types in the samples. Any that comprised
< 1% of the total amount were grouped into an "other" category for the pie charts.
The full set and values can be found in the table below.  
The plots and table show total summed values across all replicates.

```{r, pie_processing}
pie_data <- totals_data %>%
  select(pie_class, percent, condition) %>%
  group_by(pie_class, condition) %>%
  summarise(total_pie_percent = sum(percent)) %>%
  ungroup()

other_pie_data <- totals_data %>%
  filter(pie_class == "other") %>%
  select(class, percent, condition) %>%
  group_by(class, condition) %>%
  summarise(total_pie_percent = sum(percent)) %>%
  ungroup() %>%
  rename(pie_class = class)

plot_base_pie <- function(pie_condition){
  
  pie_condition_data <- pie_data %>%
    filter(condition == pie_condition)
  
  pie(
    pie_condition_data$total_pie_percent, 
    labels = pie_condition_data$pie_class, 
    col    = rainbow(nrow(pie_condition_data)),
    main   = pie_condition
  )
}

plot_base_pie <- function(pie_condition, pie_dataset){
  
  pie_condition_data <- pie_dataset %>%
    filter(condition == pie_condition)
  
  pie(
    pie_condition_data$total_pie_percent, 
    labels = pie_condition_data$pie_class, 
    col    = rainbow(nrow(pie_condition_data)),
    main   = pie_condition
  )
}
```

```{r, main_pie_charts, out.width=c('50%', '50%'), fig.show='hold', fig.height = 7}
purrr::walk(levels(pie_data$condition), plot_base_pie, pie_data)
```

### Lipids in "other" category 
```{r, main_pie_charts, out.width=c('50%', '50%'), fig.show='hold', fig.height = 7}
purrr::walk(levels(other_pie_data$condition), plot_base_pie, other_pie_data)
```



```{r, out.width=c('50%', '50%'), fig.show='hold', fig.height = 8, eval = FALSE}
other_pie <- totals_data %>%
  filter(pie_class == "other") %>%
  select(condition, class, mean_total)
  
other_pie_data <- other_pie  %>%  
  group_by(condition) %>%
  summarise(total = sum(mean_total)) %>%
  ungroup() %>%
  left_join(other_pie) %>%
  mutate(percent = (mean_total/total)*100) %>%
  group_by(class) %>%
  mutate(max_percent = max(percent)) %>%
  ungroup()

other_pie_data <- other_pie_data %>%
  select(class, percent, condition) %>%
  group_by(class, condition) %>%
  summarise(total_pie_percent = sum(percent)) %>%
  ungroup()

n_col <- n_distinct(select(other_pie_data, class))

for (this_condition in unique(other_pie_data$condition)){
  #print(this_condition)
  one_pie <- filter(other_pie_data, condition == this_condition)
  pie(
    one_pie$total_pie_percent, 
    labels = one_pie$class, 
    col = rainbow(n_col),
    main = this_condition
  )
}  

```
### Total amounts
```{r}
totals_table <- totals_data %>%
  dplyr::arrange(desc(max_percent)) %>%
  select(class, condition, sum_total, percent) %>%
  tidyr::pivot_wider(names_from = condition, values_from = c(sum_total, percent))

totals_table %>%
  DT::datatable(
      rownames = FALSE, 
      options = list(pageLength = 10, dom = "tlip"), 
     # filter = list(position = "top"),
      escape = FALSE
      ) %>%
    DT::formatRound(digits = 2, 2:7) 
```

```{r}
out_file <- paste0(params$output_folder, "/totals_table.csv") 
readr::write_csv(totals_table, file = out_file) 
```


<br><br>
<hr>
<br><br>

### Linear values by class

Bar shows median value per class. Each point is a value for an individual lipid in the class.

T-tests were performed on log transformed data. Bars are annotated with a star if there was a significant difference between the conditions (adjusted p-value < 0.05). 

```{r barplot_totals_median, fig.width = 10, fig.height=8}
pvals_by_class <- select(t_test_by_class, class, padj)

bar_data_totals <- tidy_data %>%
  dplyr::group_by(class, condition) %>%
  dplyr::summarise(median = median(value)) %>%
  ungroup() %>%
  group_by(class) %>%
  mutate(max_median = max(median)) %>%
  ungroup() %>%
  left_join(tidy_data) %>%
  select(class, value, median, condition, max_median) 

bar_data_totals <- bar_data_totals %>%
  left_join(superclasses) %>%
  group_by(superclass) %>%
  mutate(max_super_value = max(max_median)) %>%
  ungroup() %>%
  left_join(pvals_by_class) %>%
  mutate(bar_label = if_else(padj < 0.05, "*", "")) 

bar_data_totals %>%  
    ggplot(aes(x = class, y = value, fill = condition)) +
    geom_bar(aes(y = median), 
           colour = "black", 
           position= position_dodge(width = dodge_width),
           width = 0.8,
           stat="identity",
           lwd = 0.8) + 
    geom_text(aes(y = max_median), 
            nudge_y = bar_data_totals$max_super_value*0.05, 
            label = bar_data_totals$bar_label,
            size = 6,
            col = "red") +
    geom_point(position = position_jitterdodge(dodge.width=0.9, jitter.width = 0.3), shape = 21, alpha = 0.5) +
    scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, face = "bold")) +
    xlab("") +
    ylab(params$bar_class_ylabel) +
    facet_wrap(~ superclass, scales = "free")

```
<br>

### Linear values by class
  
The boxplots show summary statistics for the lipid class (the median, two hinges and two whiskers), and all outlying points individually.
Each data point for the class is overlaid.

```{r, boxplot_linear, fig.width = 10, fig.height=8}
boxplot_data <- tidy_data %>%
  left_join(superclasses) %>%
  left_join(pvals_by_class) %>%
  mutate(class_label = if_else(padj < 0.05, "*", "")) %>%
  group_by(class) %>%
  mutate(max_class_value = max(value)) %>%
  mutate(max_class_log2_value = max(log2_value)) %>%
  ungroup()

boxplot_data %>%
    ggplot(aes(x = class, y = value, fill = condition, label = class_label)) + 
    stat_boxplot(
      position= position_dodge(width = 0.9),
      width = 0.6,
      lwd = 0.8
    ) + 
    geom_text(aes(y = max_class_value), 
        size = 6,
        col = "red") +
    geom_point(position = position_jitterdodge(dodge.width=0.9, jitter.width = 0.3), shape = 21) +
    scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, face = "bold")) +
    facet_wrap(~ superclass, scales = "free")
```
<br>

### Log2 transformed values  
  
The data has been log transformed as lipid data tends to be log-normal. 
The boxplots show summary statistics for the lipid class (the median, two hinges and two whiskers), and all outlying points individually.
Each data point for the class is overlaid.

```{r, boxplot_totals, fig.width = 10, fig.height=8}
boxplot_data %>%
    ggplot(aes(x = class, y = log2_value, fill = condition, label = class_label)) + 
    stat_boxplot(
      position= position_dodge(width = 0.9),
      width = 0.6,
      lwd = 0.8
    ) + 
    geom_text(aes(y = max_class_log2_value), 
        size = 6,
        col = "red") +
    geom_point(position = position_jitterdodge(dodge.width=0.9, jitter.width = 0.3), shape = 21) +
    scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, face = "bold")) +
    facet_wrap(~ superclass, scales = "free")
```
<br>

These plots show log2 ratios between the 2 conditions, the red point shows the ratio while the lines show the confidence interval. If the confidence interval does not span 0, the difference between the 2 conditions for that particular species will be significant.
Red points to the left of the line indicate higher average levels of the lipid class in the `r exp_conditions[1]` condition, while points to the right of the line indicate higher average levels in the `r exp_conditions[2]` condition.

```{r forest_class, fig.width = 10, fig.height=8}
t_test_by_class %>%
  left_join(superclasses) %>%
  ggplot(aes(ratio, class)) +
    geom_errorbar(aes(xmin=lower.limit, xmax=upper.limit), width = 0.5, lwd=0.8, show.legend=FALSE)+
    geom_point(size=3, colour="darkred")+
    geom_vline(xintercept=0, linetype="dashed", color = "red")+
    scale_x_continuous(breaks=seq(from =-15, by=1, to=5)) +
    facet_wrap(~ superclass, scales = "free")
```


```{r, results="asis"}
# doesn't work properly using headers as it messes up the tabs
large_text <- function(x, size = 16) {
  sprintf("<span style='font-size: %spx;'>%s</span>", size, x)
}
```


<br>
<br>
<hr class="style2">  
<br>
<br>  
  
###  Volcano plot

<div class = "row">
<div class = "col-md-4">
<br><br>
The y axis shows the –log10(padj); that is the negative of the log-transformed p-values, 
adjusted for multiple comparisons.  
The x axis shows the magnitude of difference between the conditions for that lipid, 
expressed as a ratio.  
Any points to the left of the grey dashed line have higher values in `r exp_conditions[1]` compared to `r exp_conditions[2]`, and vice versa for points to the right of the grey line.  

Each point on the plot is a lipid subclass.   
Hover over a point to see the name.

The dashed red line shows significance (adj p<0.05) i.e. any points above it represent 
a significant difference between the 2 conditions.


</div>
<div class = "col-md-8">
```{r, volcano, fig.height=5, fig.width=6}
p <- t_test_res %>%
  left_join(superclasses) %>%
  ggplot(aes(x = ratio, y = -log10(padj), col = superclass, annotation = lipid_name)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), col = "red", linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
  theme(legend.title = element_blank())

plotly::ggplotly(p, tooltip = "lipid_name")
```
</div>
</div>

<br>
<br>
<hr class="style2">  
<br>
<br>

### Heatmaps

These heatmaps are scaled by row, to allow comparison between samples.

#### Lipids that change between conditions, with row clustering and column clustering  

This heatmap shows lipids with an adjusted p value < 0.05 from t-tests.   
Row labels are shown if there are < 70 rows. 

```{r, heatmap_processing}
var_above0 <- all_data %>%
  pivot_longer(-lipid_name) %>%
  group_by(lipid_name) %>%
  summarise(variance = var(value)) %>%
  filter(variance > 0) %>%
  select(lipid_name) %>%
  left_join(all_data)

heatmap_data <- tibble::column_to_rownames(var_above0, "lipid_name")
```  

```{r, heatmap_processing2}
# rearrange matrix so control samples come first
samples_conditions <- enframe(colnames(heatmap_data), name = NULL, value = "sample_name") %>%
  left_join(meta_type) 

control_samples <- samples_conditions %>%
  filter(condition == params$control) %>%
  pull(sample_name)

heatmap_data <- relocate(heatmap_data, control_samples)

col_annotation <- samples_conditions %>%
  tibble::column_to_rownames("sample_name") %>%
  arrange(condition)
  
#---------------------------------------------

classes <- tidy_data %>%
  select(lipid_name, class) %>%
  distinct()

# row names with their class
df_lipids <- var_above0 %>%
  select(lipid_name) %>%
  left_join(classes) 

row_annotation <- tibble::column_to_rownames(df_lipids, "lipid_name")

# Lipid classes need further sorting to create custom row annotations where only
# the class name is displayed rather than each individual lipid.
lipid_summary <- get_lipid_summary(df_lipids)
lipid_labels <- create_lipid_class_labels(lipid_summary, nrow(df_lipids))
lipid_colours <- create_lipid_colours(
  class_names = dplyr::pull(lipid_summary, class)
)
group_colours <- stats::setNames(c("blue", "darkgreen"), levels(col_annotation$condition))

annot_colours <- list(
  condition = group_colours,
  class = lipid_colours
)
```



```{r, heatmap1, fig.height = 8, fig.width = 10, results = "asis"}
sig_lipids <- t_test_res %>%
  filter(padj < 0.05) %>%
  pull(lipid_name)

if(length(sig_lipids) >= 2){
  
  heatmap_sig <- heatmap_data[sig_lipids,]
  row_annot_sig <- row_annotation[sig_lipids,]
  too_many_names <- if_else(nrow(heatmap_sig) > 70, FALSE, TRUE)
  
  heatmap0 <- pheatmap::pheatmap(
    heatmap_sig,
    scale = "row",
    annotation_row = row_annotation,
    annotation_col = col_annotation,
    annotation_colors = annot_colours, 
    show_rownames = too_many_names
  )
  
  heatmap0
} else cat("Not enough significant lipids to plot a heatmap")
```



#### All lipids, rows ordered by lipid subclass, columns ordered by condition, no clustering

```{r, heatmap1, fig.height = 8, fig.width = 10}
heatmap1 <- pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  labels_row = lipid_labels,
  fontsize_row = 8,
  annotation_colors = annot_colours#,
  #annotation_legend = FALSE
)

heatmap1
```

<br>
<br>

#### All lipids, rows ordered by lipid subclass, clustered by columns

```{r, heatmap2, fig.height = 8, fig.width = 10}
heatmap2 <- pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  cluster_rows = FALSE,
  labels_row = lipid_labels,
  fontsize_row = 8,
  annotation_colors = annot_colours#,
  #annotation_legend = FALSE
)

heatmap2

```

<br>
<br>

#### All lipids, with row clustering and column clustering

```{r, heatmap3, fig.height = 8, fig.width = 10}
too_many_names <- if_else(nrow(heatmap_data) > 70, FALSE, TRUE)

heatmap3 <- pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  annotation_colors = annot_colours, 
  show_rownames = too_many_names
)

heatmap3

```

