```{r}
# requires the all_data df
```

```{r}
totals <- tidy_data %>%
  group_by(class, condition) %>%
  summarise(total = sum(value)) %>%
  ungroup() %>%
  mutate(log2_total = log2(total)) %>%
  group_by(condition) %>%
  mutate(percent = (total/sum(total))*100) %>%
  ungroup() %>%
  group_by(class) %>%
  mutate(max_percent = max(percent)) %>%
  ungroup() %>%
  mutate(pie_class = if_else(max_percent > 1, class, "other"))

bar_totals_linear <- tidy_data %>%
  group_by(class, condition, sample_name) %>%
  dplyr::summarise(total = sum(value)) %>%
  ungroup() %>%
  group_by(class, condition) %>%
  dplyr::summarise(
    mean_total = mean(total),
    se_total = sd(total)/sqrt(dplyr::n())
  ) %>%
  ungroup()

bar_totals_log2 <- tidy_data %>%
  group_by(class, condition, sample_name) %>%
  dplyr::summarise(total = sum(log2_value)) %>%
  ungroup() %>%
  group_by(class, condition) %>%
  dplyr::summarise(
    mean_total = mean(total),
    se_total = sd(total)/sqrt(dplyr::n())
  ) %>%
  ungroup()

mean_log2_class <- tidy_data %>%
  dplyr::group_by(class, condition) %>%#, sample_name) %>%
  dplyr::summarise(
    mean = mean(log2_value), 
  ) %>%
  ungroup() %>%
  pivot_wider(names_from = condition, values_from = mean) %>%
  mutate(ratio = .data[[exp_conditions[1]]] - .data[[exp_conditions[2]]])
```

```{r}
t_test_by_class <- tidy_data %>%
  group_by(class) %>%#, condition, sample_name) %>%
  select(class, sample_name, condition, log2_value) %>%
  rstatix::t_test(log2_value ~ condition, paired = params$paired) %>%
  ungroup() %>%
  mutate(padj = p.adjust(p, method="BH")) %>%
  ungroup() %>%
  mutate(tadj = qt(padj/2, df)) %>%
  left_join(mean_log2_class) %>%
  mutate(SE = abs(ratio/tadj), lower.limit=ratio-1.96*SE, upper.limit=ratio+1.96*SE) 

```

<hr class="style2">  

### Summary of total amounts across all samples

The pie charts show the proportion of lipid types in the samples. Any that comprised
< 1% of the total amount were grouped into an "other" category for the pie charts.
The full set and values can be found in the table below.  
The plots and table show total summed values across all replicates.

```{r, pie_charts, out.width=c('50%', '50%'), fig.show='hold', fig.height = 7}
pie_data <- totals %>%
  select(pie_class, percent, condition) %>%
  group_by(pie_class, condition) %>%
  summarise(total_pie_percent = sum(percent)) %>%
  ungroup()

pie_cond1 <- pie_data %>%
  filter(condition == exp_conditions[1])

pie_cond2 <- pie_data %>%
  filter(condition == exp_conditions[2])

n_col <- nrow(pie_cond1)

pie(
  pie_cond1$total_pie_percent, 
  labels=pie_cond1$pie_class, 
  col=rainbow(n_col),
  main = exp_conditions[1]
)
pie(
  pie_cond2$total_pie_percent, 
  labels=pie_cond2$pie_class, 
  col=rainbow(n_col),
  main = exp_conditions[2]
  )
```

<div class = "row">
<div class = "col-md-7">

### Total amounts
```{r}
totals %>%
  dplyr::arrange(desc(max_percent)) %>%
  select(class, condition, total, percent) %>%
  tidyr::pivot_wider(names_from = condition, values_from = c(total, percent)) %>%
  DT::datatable(
      rownames = FALSE, 
      options = list(pageLength = 10, dom = "tlip"), 
     # filter = list(position = "top"),
      escape = FALSE
      ) %>%
    DT::formatRound(digits = 2, 2:5) 
```
</div>
<div class = "col-md-1">

</div>
<div class = "col-md-4">

### Mean of total amounts 
```{r}
bar_totals_linear %>%
  dplyr::arrange(desc(mean_total)) %>%
  select(class, condition, mean_total) %>%
  tidyr::pivot_wider(names_from = condition, values_from = c(mean_total)) %>%
  DT::datatable(
      rownames = FALSE, 
      options = list(pageLength = 10, dom = "tlip"), 
      escape = FALSE
      ) %>%
    DT::formatRound(digits = 2, 2:3) 
```
</div>
</div>

<br><br>
<hr>
<br><br>

### Mean of total amounts - linear scale

```{r, eval = FALSE}
bar_totals_linear %>%
  dplyr::arrange(desc(mean_total)) %>%
  ggplot(aes(x = class, y = mean_total, fill = condition)) + 
  geom_bar(
    colour = "black", 
    position= position_dodge(width = 0.8),
    width = 0.8,
    stat="identity",
    lwd = 0.8
  ) + 
  geom_errorbar(
      aes(ymin=mean_total-se_total, ymax=mean_total+se_total), 
      position = position_dodge(width = dodge_width), 
      width = 0.25,
      lwd = 0.8
    ) +
  scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, face = "bold")) +
  xlab("")
```

```{r, bar_totals, fig.width = 10, fig.height=8}
bar_totals_linear %>%
  left_join(superclasses) %>%
    ggplot(aes(x = class, y = mean_total, fill = condition)) + 
    geom_bar(
      colour = "black", 
      position= position_dodge(width = 0.8),
      width = 0.8,
      stat="identity",
      lwd = 0.8
    ) + 
    geom_errorbar(
        aes(ymin=mean_total-se_total, ymax=mean_total+se_total), 
        position = position_dodge(width = dodge_width), 
        width = 0.25,
        lwd = 0.8
      ) +
    scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, face = "bold")) +
    xlab("") +
    facet_wrap(~ superclass, scales = "free")
```

```{r forest_class, fig.width = 10, fig.height=8}
t_test_by_class %>%
  left_join(superclasses) %>%
  ggplot(aes(ratio, class)) +
    geom_errorbar(aes(xmin=lower.limit, xmax=upper.limit), width = 0.5, lwd=0.8, show.legend=FALSE)+
    geom_point(size=3, colour="darkred")+
    geom_vline(xintercept=0, linetype="dashed", color = "red")+
    scale_x_continuous(breaks=seq(from =-15, by=1, to=5)) +
    facet_wrap(~ superclass, scales = "free")
```

```{r}
#custom_forest(t_test_res, lipid_class, abundance_level = "high")
```

 
```{r, eval = FALSE}
### Totals - log2 scale  
bar_totals_linear %>%
  ggplot(aes(x = class, y = log2(mean_total), fill = condition)) + 
  geom_bar(
    colour = "black", 
    position= position_dodge(width = 0.8),
    width = 0.8,
    stat="identity",
    lwd = 0.8
  ) + 
  # geom_errorbar(
  #     aes(ymin=log2(mean_total)-log2(se_total), ymax=log2(mean_total)+log2(se_total)), 
  #     position = position_dodge(width = dodge_width), 
  #     width = 0.25,
  #     lwd = 0.8
  #   ) +
  scale_fill_manual(values = c("#9ac4d6", "#f7f7c3")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, face = "bold")) +
  xlab("")
```

```{r, results="asis"}
# doesn't work properly using headers as it messes up the tabs
large_text <- function(x, size = 16) {
  sprintf("<span style='font-size: %spx;'>%s</span>", size, x)
}
```

```{r, heatmap_processing}
var_above0 <- all_data %>%
  pivot_longer(-lipid_name) %>%
  group_by(lipid_name) %>%
  summarise(variance = var(value)) %>%
  filter(variance > 0) %>%
  select(lipid_name) %>%
  left_join(all_data)

heatmap_data <- tibble::column_to_rownames(var_above0, "lipid_name")
```  

```{r, heatmap_processing2}
classes <- tidy_data %>%
  select(lipid_name, class) %>%
  distinct()

# row names with their class
df_lipids <- var_above0 %>%
  select(lipid_name) %>%
  left_join(classes) 

row_annotation <- tibble::column_to_rownames(df_lipids, "lipid_name")

col_annotation <- enframe(colnames(heatmap_data), name = NULL, value = "sample_name") %>%
  left_join(meta_type) %>%
  tibble::column_to_rownames("sample_name")

# Lipid classes need further sorting to create custom row annotations where only
# the class name is displayed rather than each individual lipid.
lipid_summary <- get_lipid_summary(df_lipids)
lipid_labels <- create_lipid_class_labels(lipid_summary, nrow(df_lipids))
lipid_colours <- create_lipid_colours(
  class_names = dplyr::pull(lipid_summary, class)
)
group_colours <- stats::setNames(c("blue", "darkgreen"), exp_conditions)

annot_colours <- list(
  condition = group_colours,
  class = lipid_colours
)
```

<br>
<br>
<hr class="style2">  
<br>
<br>  
  
###  Volcano plot

<div class = "row">
<div class = "col-md-4">
<br><br><br><br>
The y axis shows the –log10(padj); that is the negative of the log-transformed p-values, 
adjusted for multiple comparisons.
The x axis shows the magnitude of difference between the conditions for that lipid, 
expressed as a ratio.

Each point on the plot is a lipid (subclass?).   
Hover over a point to see the name.

The dashed red line shows significance (p<0.05) i.e. any points above it represent 
a significant difference between the 2 conditions.

</div>
<div class = "col-md-8">
```{r, volcano, fig.height=5, fig.width=6}
p <- t_test_res %>%
  left_join(superclasses) %>%
  ggplot(aes(x = ratio, y = -log10(padj), col = superclass, annotation = lipid_name)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), col = "red", linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed", col = "grey") +
  theme(legend.title = element_blank())

plotly::ggplotly(p, tooltip = "lipid_name")
```
</div>
</div>

<br>
<br>
<hr class="style2">  
<br>
<br>

### Heatmaps

#### Rows ordered by lipid subclass, columns ordered by condition, no clustering

```{r, heatmap1, fig.height = 8, fig.width = 10}
heatmap1 <- pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  labels_row = lipid_labels,
  fontsize_row = 8,
  annotation_colors = annot_colours#,
  #annotation_legend = FALSE
)

heatmap1
```

<br>
<br>

#### Rows ordered by lipid subclass, clustered by columns

```{r, heatmap2, fig.height = 8, fig.width = 10}
heatmap2 <- pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  cluster_rows = FALSE,
  labels_row = lipid_labels,
  fontsize_row = 8,
  annotation_colors = annot_colours#,
  #annotation_legend = FALSE
)

heatmap2

```

<br>
<br>

#### With row clustering and column clustering

```{r, heatmap3, fig.height = 8, fig.width = 10}
too_many_names <- if_else(nrow(heatmap_data) > 70, FALSE, TRUE)

heatmap3 <- pheatmap::pheatmap(
  heatmap_data,
  scale = "row",
  annotation_row = row_annotation,
  annotation_col = col_annotation,
  annotation_colors = annot_colours, 
  show_rownames = too_many_names
)

heatmap3

```

```{r, eval = FALSE}
pdf(file = paste0(params$output_folder, "/heatmap1.pdf"))
heatmap1
dev.off()

pdf(file = paste0(params$output_folder, "/heatmap2.pdf"))
heatmap2
dev.off()

pdf(file = paste0(params$output_folder, "/heatmap3.pdf"))
heatmap3
dev.off()
```
